Clustering and Parameter Estimation for the Popularity Adjusted Block Model

### 2020-07-12 summary.Rmd
still working on writeup

---
title: "Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph"
# output: pdf_document
output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
---

@@

```{r include = FALSE}
embedding <- function(A, p = NULL, q = NULL,
                      eps = 1e-6) {
  eigen.A <- eigen(A, symmetric = TRUE)
  if (is.null(p) | is.null(q)) {
    keep <- (abs(eigen.A$values) > eps)
  } else {
    keep <- c(seq(p), seq(n, n - q + 1))
  }

  U <- eigen.A$vectors[, keep]
  S <- diag(sqrt(abs(eigen.A$values[keep])))
  return(U %*% S)
}

draw.graph <- function(P) {
  A <- apply(P, 1:2, function(p) rbinom(1, 1, p))
  A[lower.tri(A)] <- 0
  diag(A) <- 0
  A <- A + t(A)
  return(A)
}

rot.mat.2 <- function(angle, axis = 1) {
  # rotation matrix for K=2

  if (axis %in% c(1, 3)) {
    rotation <- matrix(c(cos(angle), -sin(angle), sin(angle), cos(angle)),
                       nrow = 2, ncol = 2, byrow = TRUE)
    if (axis == 1) {
      return(as.matrix(Matrix::bdiag(1, rotation, -1)))
    } else {
      return(as.matrix(Matrix::bdiag(rotation, 1, -1)))
    }
  } else if (axis == 2) {
    return(matrix(c(cos(angle), 0, sin(angle), 0,
                    0, 1, 0, 0,
                    -sin(angle), 0, cos(angle), 0,
                    0, 0, 0, -1),
                  nrow = 4, ncol = 4,
                  byrow = TRUE))
  } else {
    stop(simpleError('axis must be 1, 2, or 3'))
  }
}

hyp.rot.mat.2 <- function(angle, axis = 1) {
  # hyperbolic rotation matrix for $K=2

  if (axis == 1) {
    return(matrix(c(1, 0, 0, 0,
                    0, 1, 0, 0, 
                    0, 0, cosh(angle), sinh(angle),
                    0, 0, sinh(angle), cosh(angle)), 
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else if (axis == 2) {
    return(matrix(c(1, 0, 0, 0,
                    0, cosh(angle), 0, sinh(angle),
                    0, 0, 1, 0,
                    0, sinh(angle), 0, cosh(angle)),
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else if (axis == 3) {
    return(matrix(c(cosh(angle), 0, 0, sinh(angle),
                    0, 1, 0, 0, 
                    0, 0, 1, 0, 
                    sinh(angle), 0, 0, cosh(angle)),
                  byrow = TRUE,
                  nrow = 4, ncol = 4))
  } else {
    stop(simpleError('axis must be 1, 2, or 3'))
  }
}

construct.Q <- function(angles) {
  rot.angles <- angles[1:3]
  hyp.angles <- c(angles[1], angles[4], -angles[4])
  rot.matrices <- lapply(seq(3), function(i) rot.mat.2(rot.angles[i], i))
  hyp.matrices <- lapply(seq(3), function(i) hyp.rot.mat.2(hyp.angles[i], i))
  matrices <- c(rot.matrices, hyp.matrices)
  return(Reduce(`%*%`, matrices))
}

obj.fun <- function(angles, Z.hat) {
  R <- construct.Q(angles)

  XU.c <- Z.hat %*% R
  # clusters <- apply(XU.c[, 1:2] ** 2, 1, which.min)
  clusters <- c(rep(1, 32), rep(2, 32))
  in.prods <- XU.c[clusters == 1, ] %*% t(XU.c[clusters == 2, , drop = FALSE])
  return(
    sum(apply(XU.c[, 1:2] ** 2, 1, min)) +
      sum(apply(cbind((XU.c[, 3] - XU.c[, 4]) ** 2,
                      (XU.c[, 3] + XU.c[, 4]) ** 2),
                1, min))
    )
  # return(
  #   sum(apply(XU.c[, 1:2] ** 2, 1, min)) +
  #     Matrix::norm(in.prods, 'F') +
  #     sum((XU.c[clusters == 1, 3] - XU.c[clusters == 1, 4]) ** 2) +
  #     sum((XU.c[clusters == 2, 3] + XU.c[clusters == 2, 4]) ** 2)
  # )
}

transform.embedding <- function(Z.hat) {
  init.angles <- runif(4, -pi, pi)
  angles.1 <- optim(init.angles, function(x) obj.fun(x, Z.hat), 
                    method = 'SANN')$par
  angles.2 <- optim(angles.1, function(x) obj.fun(x, Z.hat),
                    method = 'Nelder-Mead',
                    control = list(abstol = 1e-10, maxit = 1e3))$par
  R <- construct.Q(angles.2)
  return(Z.hat %*% R)
}
```

@@

### The popularity adjusted block model (PABM)[^pabm]

@@

### The generalized random dot product graph (GRDPG)[^grdpg]

@@

### Connecting the PABM to the GRDPG for $K = 2$

@@

### Generalization to $K > 2$

@@

#### Example: $K = 3$

@@

Then $Y = X \Pi$ where 
$\Pi = \begin{bmatrix} 
1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1
\end{bmatrix}$

@@

#### Additional notes and observations

* $XU$ is a convenient but not necessarily unique characterization of the 
latent positions. This comes into play when we try to estimate the latent 
positions---the estimate is only unique up to matrix multiplication by an 
arbitrary matrix in $\mathbb{O}(p, q)$.
    * i.e., let $Q \in \mathbb{O}(p, q)$, and let $P = (XU) I_{pq} (XU)^\top$. 
    Then we can also decompose $P = (XUQ) I_{pq} (XUQ)^\top$
    $= (XU) (Q I_{pq} Q^\top) (XU)^\top$
    $= (XU) I_{pq} (XU)^\top = P$
    * $\mathbb{O}(p, q) = 
    \{Q \in \mathbb{R}^{(p+q) \times (p+q)} : Q I_{pq} Q^\top = I_{pq}\}$
* If we use the latent positions described by $XU$, the dot product of two 
points in different clusters equals 0. Points that are not within the same 
cluster are orthogonal to each other (again, if we take $XU$ as the true 
latent positions).
* Even though the latent positions $XU$ follow these conditions, we can 
multiply by any matrix $Q \in \mathbb{O}(p, q)$ and have these conditions 
fail.

### Recovery/Estimation/Inference

The main goals are:

1. Determining the underlying cluster/community memberships/structure
2. Estimating the latent positions $XU$
3. Estimating $\lambda^{(kl)}$ for $k, l = 1, ..., K$

#### Previous results

Theorem 5 from the GRDPG paper[^grdpg] says the following:

* Let 
    * $X$ be an latent position matrix with latent positions as $n$ rows 
    $x_i^\top$ in $d$ dimensions
    * $A \sim GRDPG_{p, q}(X)$
    * $\hat{x}_i$'s be the adjacency embedding of $A$
* Then 
    * $\max\limits_{i = 1, ..., n} || Q \hat{x}_i - x_i || = 
    O_P(\frac{(\log n)^c}{n^{1/2}})$
    * $Q \in \mathbb{O}(p, q)$ (unidentifiable)
    * $c > 0$

Under some sparsity conditions.

So we can only identify the embedding up to matrix multiplication.

Both the GRDPG paper and the PABM paper[^pabm] starting with the spectral 
embedding (adjacency or Laplacian) of $A$ and performing a clustering step 
($K$-means, GMM, etc.). Unfortunately, there aren't any theoretical guarantees 
for the clustering step.

The PABM paper suggests subspace clustering on the embedding, and we can 
clearly see the justification by looking at $XU$. The rows of $XU$ that are 
within the same cluster lie in a $K$-dimensional subspace. We can further 
see that the $K$ subspaces defined by the $K$ clusters are all orthogonal to 
each other. However, in practice, the unidentifiable $Q$ matrix breaks this 
property when we decompose $A$.

***TODO: Investigate CLT***

#### Proposed Methods

Let $z_i^\top$ denote the $i$^th^ row of $XU$.  
Let $\gamma(x, y)$ denote the cosine similarity between $x$ and $y$.  
Then $\gamma(z_i, z_j) = 0$ if vertices $i$ and $j$ belong to different 
clusters, and $\gamma(z_i, z_j) > 0$ almost surely if $i$ and $j$ belong to the 
same cluster.

This then gives us the result:

* Let $\hat{Z}$ denote the adjacency embedding of 
$A \sim PABM(\{\lambda^{(kl)}\}_K)$ with rows $\hat{z}_i^\top$.
* Then $\exists Q \in \mathbb{O}(p, q)$ s.t. 
$\max\limits_{i, j = 1, ...,n} | \gamma(Q \hat{z}_i, Q \hat{z}_j) | =
O_P(\frac{(\log n)^c}{n^{1/2}})$ if vertices $i$ and $j$ are in different 
clusters.[^convergence]
* We can state a similar result using the Laplacian embedding of $A$.

[^convergence]: TODO: Check that the convergence rate is true (I'm pretty sure 
it is since we are taking the dot product of two vectors and then dividing 
by their norms, so we should get the same convergence rate as in the GRDPG 
result). Or alternatively, 
$|(Q \hat{z}_i)^\top (Q \hat{z}_j)| = O_P(\frac{(\log n)^{2c}}{n})$.

***TODO: CLT results***

Thus if we can recover $Q$, we can assign cluster memberships based on the behavior of the cosine similarity matrix.[^clt] The problem here is that we cannot get rid of the intractable matrix $Q$.

Also note that $Q$ is not merely a rotation matrix. Multiplying by $Q$ changes 
inter-point distances (and angles).

[^clt]: TODO: Use CLT to describe deviation from 1 or 0, e.g., if $i$, $j$ in 
different clusters then what is the distribution/density of 
$\gamma(\hat{z}_i, \hat{z}_j)$, which should be centered at 0.

However, perhaps we can use the properties of $XU$ to estimate $Q$. $XU$ 
consists of points on $K$ $K$-dimensional orthogonal hyperplanes. The PABM 
paper uses the hyperplanes part to justify subspace clustering but does not 
use the fact that they are orthogonal hyperplanes---the orthogonality is 
broken by the decomposition of $P$. If $Z$ is the adjacency embedding of $P$, 
then $Z = X U Q^{-1}$ for some unknown $Q \in \mathbb{O}(p, q)$. However, 
the multiplication by $Q^{-1}$ does not affect the fact that the points lie 
along (not necessarily orthogonal) hyperplanes, which is a justification for 
using subspace clustering to identify the community memberships.

For now, we will focus on he case $K = 2$. In this scenario, we have the 
following: We wish to find $Q \in \mathbb{O}(3, 1)$ such that $Z Q$ "looks like" 
$XU$. $X U = \begin{bmatrix} 
\lambda^{(11)} & 0 & \lambda^{(12)} / \sqrt{2} & \lambda^{(12)} / \sqrt{2} \\
0 & \lambda^{(22)} & \lambda^{(21)} / \sqrt{2} & -\lambda^{(21)} / \sqrt{2}
\end{bmatrix}$, and we can observe the following properties:

1. If two rows of $XU$ are not in the same cluster, they are orthogonal.
2. If vertex $i$ is in the first cluster, its second entry is $0$. Likewise,
if vertex $j$ is in the second cluster, its first entry is $0$. 
3. If vertex $i$ is in the first cluster, its third and fourth entries are 
identical. Similarly, if vertex $j$ is in the second cluster, the sum of its 
third and fourth entries is equal to $0$.

It is straightforward to generalize these properties to $K > 2$.

So we need to find $Q \in \mathbb{O}(3, 1)$ such that $\hat{Z} Q$ meets these 
criteria as closely as possible. One optimization problem that could arise 
from this is:

$$\begin{split}
  \arg\min_Q & ||\gamma^{(1)}||^2 + ||\gamma^{(2)}||^2 \\
  \text{s.t. } & \Xi = \hat{Z} Q \text{ with rows } \xi_i^\top \\
  & \gamma^{(1)}_i = \min (|\xi_{i1}|, |\xi_{i2}|) \\
  & \gamma^{(2)}_i = \min(|\xi_{i3} + \xi_{i4}|, |\xi_{i3} - \xi_{i4}|) \\
  & Q \in \mathbb{O}(3, 1)
\end{split}$$

Again, this part is easily generalizable to $K > 2$.

If we embed $P$ instead of $A$, the optimal value of this objective is $0$.

A possible pseudo-algorithm for this is as follows:

1. Start with affinity matrix $A$ and its adjacency embedding $\hat{Z}$
2. Pick some $\tilde{Q} \in \mathbb{O}(3, 1)$.
3. Compute $\Xi = \hat{Z} \tilde{Q}$ and vectors $\gamma^{(1)}$, $\gamma^{(2)}$.
4. Compute the objective function.
5. Pick a new $\tilde{Q} \in \mathbb{O}(3, 1)$ that lowers the objective 
function, and repeat steps 3-5.

Note that this doesn't include orthogonality between points in different 
clusters, but the other two criteria imply orthogonality.

The advantage of this approach over others is that we can directly estimate 
$X U$ and therefore $X$ (since we know what $U$ is exactly for any given $K$) 
instead of relying on subspace clustering. Once we have the embedding 
$\hat{Z}$ and we find $\hat{Q}$, we can set $\hat{X} = \hat{Z} \hat{Q} U^\top$, 
and the elements of $\hat{X}$ will correspond to the PABM vectors 
$\lambda^{(kl)}$ (exactly if the embedding is of $P$, approximately if the 
embedding is of $A$).

The difficulty is then in choosing $Q$ from $\mathbb{O}(3, 1)$. It turns out 
that $\mathbb{O}(3, 1)$ is the Lorentz group[^lorentz] and matrices of this 
group are equivalent to Lorentz transformations, which describe the warping of 
spacetime in special relativity[^convention]. Any matrix 
$Q \in \mathbb{O}(3, 1)$ can be expressed as a product of six matrices that 
depend on four parameters[^generator], i.e., 
$Q = R_1(\theta) R_2(\phi) R_3(\psi) S_1(\theta) S_2(\tau) S_3(-\tau)$.The six 
matrices are analogous to three rotation matrices and six squeeze mappings.

$R_1(\theta) = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & \cos \theta & \sin \theta & 0 \\
0 & -\sin \theta & \cos \theta & 0 \\
0 & 0 & 0 & -1
\end{bmatrix}$

$R_2(\phi) = \begin{bmatrix}
\cos \phi & 0 & \sin \phi & 0 \\
0 & 1 & 0 & 0 \\
-\sin \phi & 0 & \cos \phi & 0 \\
0 & 0 & 0 & -1
\end{bmatrix}$

$R_3(\psi) = \begin{bmatrix}
\cos \psi & \sin \psi & 0 & 0 \\
-\sin \psi & \cos \psi & 0 & 0 \\
0 & 0 & 1 & 0 \\
0 & 0 & 0 & 1
\end{bmatrix}$

$S_1(\theta) = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 \\
0 & 0 & \cosh \theta & \sinh \theta \\
0 & 0 & \sinh \theta & \cosh \theta
\end{bmatrix}$

$S_2(\tau) = \begin{bmatrix}
1 & 0 & 0 & 0 \\
0 & \cosh \tau & 0 & \sinh \tau \\
0 & 0 & 1 & 0 \\
0 & \sinh \tau & 0 & \cosh \tau
\end{bmatrix}$

$S_3(-\tau) = \begin{bmatrix}
\cosh \tau & 0 & 0 & -\sinh \tau \\
0 & 1 & 0 & 0 \\
0 & 0 & 1 & 0 \\
-\sinh \tau & 0 & 0 & \cosh \tau
\end{bmatrix}$

We can write $Q = Q(\theta, \phi, \psi, \tau)$ and tweak the four angles until 
convergence.

**TODO: Show that for arbitrary $K$, we need $K^2$ angles and $K (K+1)$ 
matrices.**

**TODO: Compute gradients(?)**

[^lorentz]: https://en.wikipedia.org/wiki/Lorentz_group

[^convention]: There is actually a multiplication by $-1$ to account for here.

[^generator]: https://iopscience.iop.org/book/978-1-6817-4254-0/chapter/bk978-1-6817-4254-0ch1

#### Alternative approaches

1. If we know the cluster memberships (or if we estimate them using subspace 
clustering), then we can write a perhaps easier or more robust optimization 
problem:  
Rearrange $\hat{Z}$ such that the first $n_1$ rows are in cluster 1 and the 
next $n_2$ rows are in cluster 2. Then solve the following.
$$\begin{split}
\arg\min_Q & ||\Xi_{1:n_1, 1}||^2 + ||\Xi_{n_1+1 : n, 2}||^2 + 
||\Xi_{1:n_1, 3} - \Xi_{1:n1, 4}||^2 + 
||\xi_{n_1+1 : n, 3} + \xi_{n_1 + 1, 4}||^2 \\
\text{s.t. } & \Xi = \hat{Z} Q \\
& Q \in \mathbb{O}(3, 1)
\end{split}$$  
Numerical expeirments thus far suggest that this method does not perform any 
better than estimating $X U$ directly without a clustering step (although 
more investigation is required to say anything for sure).

2. $Q \in \mathbb{O}(3, 1)$ can be written as:  
$Q = \begin{bmatrix} M & -a \\ -b^\top & \gamma \end{bmatrix}$[^decomp]. This 
might produce a better parameterization of the $Q$ matrices, although so far I 
haven't had much luck here. We can see that if $Q I_{3, 1} Q^\top = I_{3, 1}$, 
then we get the following constraints:
    * $M M^\top - a a^\top = I$
    * $\gamma a = M b$
    * $\gamma^2 = b^\top b + 1$

[^decomp]: https://en.wikipedia.org/wiki/Lorentz_transformation

### Numerical example

#### PABM with $K = 2$

We will draw the elements of $\lambda^{(11)}$ and $\lambda^{(22)}$ 
independently  from a beta distribution with shape parameters $a$ and $b$ such 
that $a > b$. 
The elements of $\lambda^{(12)}$ and $\lambda^{(21)}$ will also be 
independently drawn from a beta distribution, this time with shape parameters 
$c$ and $d$ such that $c < d$. This ensures that it is more likely for a pair 
of vertices from the same cluster to be connected than it is for a pair of 
vertices from different clusters.

```{r}
# parameters
n1 <- 2 ** 5
n2 <- n1
n <- n1 + n2
Ipq <- diag(c(1, 1, 1, -1))

# cluster assignments
z <- c(rep(1, n1), 
       rep(2, n2))

# construct P
lambda11 <- rbeta(n1, 3, 1)
lambda22 <- rbeta(n2, 3, 1)
lambda12 <- rbeta(n1, 1, 3)
lambda21 <- rbeta(n2, 1, 3)
X <- cbind(c(lambda11, rep(0, n2)),
           c(lambda12, rep(0, n2)),
           c(rep(0, n1), lambda21),
           c(rep(0, n1), lambda22))
Y <- cbind(c(lambda11, rep(0, n2)),
           c(rep(0, n1), lambda21),
           c(lambda12, rep(0, n2)),
           c(rep(0, n1), lambda22))
P <- X %*% t(Y)

# draw affinity matrix
A <- draw.graph(P)

# U for K=2
U <- cbind(c(1, 0, 0, 0),
           c(0, 0, 0, 1),
           c(0, 1 / sqrt(2), 1 / sqrt(2), 0),
           c(0, 1 / sqrt(2), -1 / sqrt(2), 0))

# latent positions
XU <- X %*% U

# print(XU)

Z <- embedding(P)
Zhat <- embedding(A, 3, 1)
```

Probability matrix:

```{r, fig.width = 3}
plot(P, border = NA)
```

One realization $A$ of $P$:

```{r}
qgraph::qgraph(A, groups = factor(z))
```

```{r fig.width = 3}
plot(A, border = NA)
```

The "true" latent positions $X U$:

```{r fig.width = 4}
pairs(XU, col = z, asp = 1)
```

$Z$, the adjacency embedding of $P$:

```{r fig.width = 4}
pairs(Z, asp = 1, col = z)
```

$\hat{Z}$, the adjacency embedding of $A$:

```{r fig.width = 4}
pairs(Zhat, asp = 1, col = z)
```

Recovering $X U$ from $Z$:

```{r cache = FALSE, fig.width = 4}
XU.prime <- transform.embedding(Z)
pairs(XU.prime, asp = 1, col = z)

# XU.prime
```

Recovering $X U$ from $\hat{Z}$:

```{r cache = TRUE, fig.width = 4}
XU.hat <- transform.embedding(Zhat)
pairs(XU.hat, asp = 1, col = z)

# XU.hat
```

Some brief checking shows that we are able to recover the $XU$ and the 
$\lambda^{(kl)}$'s from $P$ exactly, which suggests that this method should 
work asymptotically, but it is not as well behaved when we use $A$.

### Additional TODOs

* We are currently treating the optimization step as a black box (just using 
Nelder-Mead or simulated annealing at the moment). Since 
$Q = Q(\theta, \phi, \psi, \tau)$ and is smooth with respect to each of those 
four parameters, we should be able to use a more disciplined 
gradient/Hessian-based approach.
* Doing the above should provide us a good way to describe the convergence 
behavior of estimating $Q$.
* Is the connection between PABM and special relativity a happy coincidence, or 
is there something more to explore here?
* Generalize optimization step to $K > 2$.
* The estimate $\hat{Z} \hat{Q}$ doesn't necessarily obey GRDPG rules. How 
can we force it to?
* What happens when we let $\lambda^{(kl)}$'s be random variables instead of 
fixed vectors, e.g., $\lambda^{(kl)}_i \sim Beta(a_{kl}, b_{kl})$ (or some 
Dirichlet equivalent)?

>>>>

---
title: "Clustering and Parameter Estimation for the Popularity Adjusted Block Model"
# output: pdf_document
output: html_document
# geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
urlcolor: blue
header-includes:
- \usepackage{float}
- \usepackage{mathtools}
abstract: |
  this
  is
  an
  abstract
---

@@

@@

# 1 Introduction

## 1.1 Previous work

# 2 Connecting the Popularity Adjusted Block Model to the Generalized Random Dot Product Graph

## 2.1 The popularity adjusted block model (PABM)[^pabm]

@@

## 2.2 The generalized random dot product graph (GRDPG)[^grdpg]

@@

## 2.3 Connecting the PABM to the GRDPG for $K = 2$

@@

## 2.4 Generalization to $K > 2$

@@

**Example: $K = 3$**

@@

Then $P = X Y$ and $Y = X \Pi$ where 
$\Pi = \begin{bmatrix} 
1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 1 & 0 \\
0 & 0 & 1 & 0 & 0 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 1
\end{bmatrix}$

@@

# 3 Methodology

Two inference objectives arise from the PABM:

1. Cluster membership identification (up to permutation).
2. Parameter estimation (estimating $\lambda^{(kl)}$'s).

Here, we will focus more on (1) and pose possible methods for (2). In our 
methods, we assume that $K$, the number of clusters, is known beforehand.

## 3.1 Cluster membership identification

### 3.1.1 Using edge probability matrix $P$

We previously stated one possible set of latent positions that result in the 
edge probability matrix of a PABM, $P = (XU) I_{pq} (XU)^\top$. If we have 
(or can estimate) $XU$ directly, then both the clustering and parameter 
identification problem are trivial since $U$ is orthonormal and fixed for each 
value of $K$. However, direct identification or estimation of $XU$ is not 
possible.

If we decompose $P = Z I_{pq} Z^\top$, then
$\exists Q \in \mathbb{O}(p, q)$ such that $XU = Z Q$. Even if we start with 
the exact edge probability matrix, we cannot recover the "original" latent 
positions $XU$. Note that unlike in the case of the regular random dot product 
graph, $Q$ is not an orthogonal matrix. If $z_i$'s are the rows of $XU$, then 
$||z_i - z_j||^2 \neq ||Q z_i - Q z_j||^2$, and 
$\langle z_i, z_j \rangle \neq \langle Q z_i, Q z_j \rangle$. This prevents 
us from using the properties of $XU$ directly. In particular, if 
$Q \in \mathbb{O}(n)$, then we could use the fact that 
$\langle z_i, z_j \rangle = \langle Q z_i, Q z_j \rangle = 0$ if vertices $i$ 
and $j$ are in different clusters.

We can note from the explicit form of $XU$ that it represents points in 
$\mathbb{R}^{K^2}$ such that points within each cluster lie on $K$-dimensional 
subspaces. Furthermore, the subspaces are orthogonal to each other. 
Multiplication by $Q \in \mathbb{O}(p, q)$ removes the orthogonality property 
but retains the property that each cluster is represented by a $K$-dimensional 
subspace. Using this property, previous work proposes the use of subspace 
clustering while acknowledging some of its shortcomings.

**Theorem 1.** 
*Let $P = V D V^\top$ be the spectral decomposition of the edge probability
matrix.*

### 3.1.2 Using adjacency matrix $A$

## 3.2 Parameter estimation

# 4 Simulated Examples
